---
title: "Process ABR Data for Deep Learning"
output:
  html_document:
    df_print: paged
---

This notebook outlines the process for preparing antibiogram data for deep learning analysis. It includes steps for data cleaning, feature extraction, and preparation of training, validation, and test datasets based on taxonomic annotation.


```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(data.table)
library(caret)
source("utils.r")
```
## Importing Data

Read in the antibiogram data from an RDS file. This dataset contains measurements from antibiogram tests. We filter for manually annotated entries to ensure data quality.

```{r}
annot <- readRDS("bacdvie_live_20230830.rds")
abr_data <- as.data.frame(annot$met_antibiogram) %>%
  filter(manual_annotation == 1)
```

## Preparing Target Data for Deep Learning

Generate a CSV file containing target values for deep learning models. This involves calculating the mean diameters for each antibiotic based on their start and end range values.


```{r}
# Generate target table with mean diameter values for antibiotics
antibiotic_columns <- grep("_rgStart|_rgEnd", names(abr_data), value = TRUE)
antibiotics <- unique(sub("_[^_]+$", "", antibiotic_columns))
target_data <- abr_data %>% select(ID_strains, all_of(antibiotic_columns))

for (antibiotic in antibiotics) {
  start_col <- paste0(antibiotic, "_rgStart")
  end_col <- paste0(antibiotic, "_rgEnd")
  target_data <- target_data %>%
    mutate(!!antibiotic := rowMeans(select(., starts_with(antibiotic)), na.rm = TRUE))
}

# Drop start and end columns, retaining mean values
target_data <- target_data %>% select(ID_strains, all_of(antibiotics))
write.csv(target_data, "target_table.csv", row.names = FALSE)
```

## Preparing Input Data

Create a CSV file for the input features. This includes environmental conditions under which each antibiogram test was conducted. Missing values are replaced with default values.


```{r}
# Select environmental conditions and replace NA values with defaults
input_table <- abr_data %>%
  select(ID_strains, Medium_antibiogram, Inc_temp_antibiogram, Inc_time_antibiogram, Inc_ox_antibiogram) %>%
  mutate(Inc_ox_antibiogram = ifelse(is.na(Inc_ox_antibiogram), 'aerob', Inc_ox_antibiogram),
         Inc_temp_antibiogram = ifelse(is.na(Inc_temp_antibiogram), 37, Inc_temp_antibiogram),
         Inc_time_antibiogram = ifelse(is.na(Inc_time_antibiogram), 1, Inc_time_antibiogram))

# Check for and warn about remaining NA values
remaining_na <- sum(is.na(input_table))
if (remaining_na > 0) {
  warning("There are still NA values remaining in the input_table.")
}

# Export input data to CSV
write.csv(input_table, "input_table.csv", row.names = FALSE)
```

## Merging Additional Metadata

Incorporate taxonomic and filename information for a comprehensive dataset. This aids in stratifying the data by taxonomic classification.

```{r}
# Read taxonomic and filename data
taxa <- readRDS("strainsData.rds")
taxa <- data.frame(ID_strains = taxa$ID_strains, genus = taxa$genus)
filenames <- readRDS("filenameData.rds")

# Merge datasets
both <- merge(target_data, filenames, by = "ID_strains") %>%
  select(filename, everything())
both2 <- merge(both, taxa, by = "ID_strains")

# Analyze genus distribution
genus_distribution <- table(both2$genus)
genus_frequency <- as.data.frame(table(both2$genus))
```

## Dataset Partitioning

Partition the data into training, validation, and test sets based on taxonomic classification, ensuring that each set includes a diverse representation of genera.

```{r}
# Partitioning logic
test_percentage <- 0.20
num_test_entries <- ceiling(nrow(both2) * test_percentage)
set.seed(123) # Ensure reproducibility

shuffled_genus_frequency <- genus_frequency[sample(nrow(genus_frequency)), ]
shuffled_genus_frequency$Var1 <- as.character(shuffled_genus_frequency$Var1)

cumulative_frequency <- 0
selected_genera <- character(0)

for (i in 1:nrow(shuffled_genus_frequency)) {
  if (cumulative_frequency < num_test_entries) {
    selected_genera <- c(selected_genera, shuffled_genus_frequency$Var1[i])
    cumulative_frequency <- cumulative_frequency + shuffled_genus_frequency$Freq[i]
  } else {
    break
  }
}

# Split data into test, training, and validation sets
test_data <- both2 %>% filter(genus %in% selected_genera)
train_val_data <- both2 %>% filter(!(genus %in% selected_genera))

trainIndex <- createDataPartition(train_val_data$genus, p = 0.7, list = FALSE, times = 1)
train_data <- train_val_data[trainIndex, ]
validation_data <- train_val_data[-trainIndex, ]

# Label dataset type
train_data$dataset_type <- 'train'
validation_data$dataset_type <- 'validation'
test_data$dataset_type <- 'test'

# Combine and clean final dataset
final_data <- rbind(train_data, validation_data, test_data)
final_data$assembly_lvl <- NULL
final_data$size <- NULL
final_data$genus <- NULL
```

## Aligning Input Table with Final Data

Ensure the input table reflects the same partitioning and order as the final dataset for consistent model training.

```{r}
# read taxonomic information for the ID_strains
# Aligning input table
both <- merge(input_table, filenames, by = "ID_strains") %>%
  select(filename, everything())
both2 <- merge(both, taxa, by = "ID_strains")

# Apply the same selection of train/test on input_table
test_data_input <- both2 %>% filter(genus %in% selected_genera)
train_val_data_input <- both2 %>% filter(!(genus %in% selected_genera))

train_val_data_input$assembly_lvl <- NULL
train_val_data_input$size <- NULL
train_val_data_input$genus <- NULL
train_val_data_input$ID_strains <- NULL

final_data_labels <- data.frame(filename = final_data$filename, labels = final_data$dataset_type)
final_data$dataset_type <- NULL
final_data$ID_strains <- NULL

```

## Export Finalized Datasets

Export the processed datasets for use in machine learning models. This includes the main dataset, labels, and input features.

```{r}
write.csv(final_data, file="training_abiogram_2024-02-21.csv", quote=FALSE, row.names=FALSE, col.names=TRUE)
write.csv(final_data_labels, file="training_abiogram_labels_2024-02-21.csv", quote=FALSE, row.names=FALSE, col.names=TRUE)
write.csv(train_val_data_input, file="training_abiogram_input_2024-02-21.csv", quote=FALSE, row.names=FALSE, col.names=TRUE)

# For a version of the input table with fewer features
train_val_data_input$Medium_antibiogram <- NULL
train_val_data_input$Inc_time_antibiogram <- NULL
write.csv(train_val_data_input, file="training_abiogram_input_smaller_2024-02-21.csv", quote=FALSE, row.names=FALSE, col.names=TRUE)
```

```{r}
head(final_data)
head(final_data_labels)
head(train_val_data_input)
```


